Language Skills in an Ethnically Segmented Labour market: Estonia 1989 -- 2012

### 2018-07-22 paper/ijm.tex
introduction, in particular CEE stuff, redone

The bulk of the existing literature analyses the effect of knowledge
of the local majority language for immigrants.  It is well known that
language skills are associated with a substantial income premium
between 5 and 35 percent
\citep{Chiswick1995,bleakley+chin2004,shields+price2002,leslie+lindley2001,chiswick+miller2002,Chiswick2003,Chiswick2010,Chiswick2015, Dustmann2003}.  For
instance, \citet{chiswick2008} argues that ``proficient'' speakers
typically earn about 15\% more than ``not proficient'' speakers.
The effect works largely through improving educational attainment
\citep{bleakley+chin2004,rooth+saarela2007native}.  Language skills are
also complementary to other types of human capital
\citep{chiswick+miller2007}.
Immigrants who are fluent in the local language also suffer
substantially less from unemployment \citep{shields+price2002, Dustmann2003}.

@@

There is also an increasing body of literature focusing on relationship
between other languages than the local main business language and
income.  These studies generally find
that English is a significant predictor of better income
\citep{Lang2009, Casale2011, Toomet2011, Williams2011, azam+2013EDandCC, isphording2013, fabo+2017E}.

The role of other languages, in particular in multi-lingual economy, is
less clear.  For instance,
\cite{Drinkwater1997} finds
a positive effect for Welsh speaking workers in
Wales and \cite{Armstrong2015} shows that bilingual native French-speakers enjoy
a significant premium both inside and outside of Quebec.
\citet{FrancoisGrin1998}, in contrast,
shows that Italian language skills in Swiss non-Italian speaking
regions are associated with lower wage.  It remains unclear if
this is a true effect or just an omitted variable bias.

Few previous studies have provided a comprehensive analysis of a labour
market from the viewpoint of language skills.
\citet{Chiswick1995, chiswick+miller2002, chiswick+miller2007, Bellante1998, Chiswick2010} analyse the effects of dominant language fluency for earnings among immigrants in the United States
\citet{Chiswick1995}  also do this for Canada, Australia and Israel.
\citet{Dustmann2003, shields+price2002} explore of immigrant assimilation into the labour market of the United Kingdom.

\citet{leslie+lindley2001} analyse both employment and income but
focus on a single year and English language only.
\citet{YaoandOurs2015} only analyse the role of Dutch language
on labour market performance in Netherlands.  
\fxwarning{ot:add CEE stuff here?}
\fxwarning{SKB: Let's collect first enough material and then decide, but I can start adding here some notes from \cite{Kahanec2009,Kahanec2010}.}
%Start here CEE stuff for now

Comparing the results for Estonia with other Central Eastern European (CEE) countries and former Soviet Union countries is rather difficult due to the lack of data about language skills.
Information about ethnicity is widely available, but no information about individual native and foreign language skills.
For example, the Latvian Labour Force Survey does not (did not) contain questions regarding language skills.
While the language policies in the Baltic states are discussed in \cite{Hogan-Brun2008,Hogan-Brun2007}, the implications for the labour market are not discussed.
\cite{Dietrich2005} discusses the status of the Russian language and the related language policy in former Soviet Union countries.
\cite{Giddings2003} discusses the ethnic wage gap in Bulgaria between ethnic Bulgarians and ethnic Turks. The wage gap increased from 0.16 in 1986 to 0.4 log points on average in 1997.
\cite{Grasmane2011} use information from a survey conducted at the Latvian University of Agriculture with undergraduate, master, doctoral students to find out  how foreign language skills impact employability.
They find that the better foreign language skills are related to higher willingness to work abroad.
\cite{Alan2015}  show that Russian language proficiency stills matters for labour market outcomes in three former Soviet Union countries (Armenia, Azerbaijan and Georgia). Russian language skills increase the likelihood of being employed between six and nine percentage points. 
Similar articles about other former Soviet Union countries are not known to the authors.
%\enquote{During the Soviet era, proficiency in the Russian language was often a ticket to
%	attractive employment opportunities in the member republics. Does it still contribute
%	to securing employment in the former Soviet republics after two decades of transition?
%	Using data from Armenia, Azerbaijan and Georgia in the years 2008–2010,
%	this paper demonstrates that Russian language skills remain economically valuable.
%	The baseline estimates suggest that Russian language skills increase probability of
%	employment by about 6 (males) and 9 (females) percentage points. Our results bear
%	important implications for the ongoing debates on language policies in the post-
%	Soviet countries.} 
\fxnote{SKB: Quite detailed article about Russian language proficiency in three former Soviet Union states. Needs further comments from us and find similar articles.}
\cite{Kahanec2009} provide an overview about the roles of foreign origin and citizenship in the labor markets of CEE Member States of the EU. 
 They use the EU Survey of Income and Living Conditions (EU-SILC) dataset to  map and compare the roles of foreign origin and citizenship for labor market performance.
Using Oaxaca and Fairlie decompositions, they show that there is native/non-native divide in the national labour markets which reduces the employment and earnings prospects of men and women with foreign origin and/or citizenship.
%They find that \enquote{both foreign origin and citizenship remain important barriers to non-native labor market integration}
For Latvia and Estonia, they find that ethnic related employment policies are most relevant for the Russian speaking minorities which are not citizens of the two states. Once employed, the negative impact of non-nativity is less important, but labor market segregation/segmentation remains important.
This is in line with our findings in this, although we focus more on the language aspect as a mediator of the ethnic dimension.
%\enquote{In Estonia and Latvia, in particular, employment
%	policies seem to be most relevant for non-immigrant non-citizens. Once employed,
%	non-nativity may play a less negative role in the labor market for some groups,
%	although labor market segmentation remains a significant issue.}
%\fxnote{\cite{Kahanec2009} essentially confirms your results in \cite{Leping2008,Toomet2011}}
\cite{Kahanec2010} discuss the difficulties of ethnic minorities in the labour market of the European Union.
Their analyses are made difficult by the lack of (enough) information about the ethnic composition in available EU-wide harmonized microdata sets, such as the
European Social Survey (ESS), Eurobarometer, EU Labour Force Survey (EU LFS), EU
Survey of Income and Living Conditions (EU SILC) or the European Community
Household Panel (ECHP). 
Still, they find confirmation of the results in \cite{Kahanec2009}, that a finding a job is difficult as a member of an ethnic minority. 
\cite{Constant2011} document the ethnic wage gap in Ukraine between ethnic  Russians and ethnic Ukrainians in favour of ethnic Russians.
%But after being in the job, the labour market prospect are somewhat similar to the ethnic majority members.
%Continue with existing stuff
In case of Estonia,
\citet{Toomet2011} shows that the majority language gives little wage
premium for minority men.  However, he analyses neither women, nor
unemployment.  \citet{Lindemann2013} shows that Estonian fluency
is associated with faster job finding rate while while it does not
explain the lower occupational prestige of Russian speakers.  However,
she only includes recent school graduates in her study.
Even earlier, \cite{Kroncke1999} found for Estonian Labour Force Survey that in 1989, there was no evidence for ethnic wage discrimination in Estonia, while in 1994 there was already a substantial wage discrimination against ethnic Russians. But in line with our findings here and in \cite{Toomet2011}, Estonian language skills did not affects wages.

In this paper, we analyse the relationship between both
majority and minority language skills, and English for both
wage and unemployment.  We analyse Estonia, a former Soviet republic
and now a rapidly growing and
transforming Easter European country.  It is characterized by a stark linguistic division
where the Russian-speaking community forms almost 30\% of
the population.  We provide long--term trends from the last years
of the planned economy in 1989 till the post-recession time in 2012.

We find that the Estonian language, the majority and the official
language of the country, is associated with very
small albeit slowly increasing wage premium over
the years.  Interestingly, this is true only for women, Russian men enjoy
virtually no premium from the majority language skills.
Russian skills, on the other hand, are clearly related to
higher wage but the relationship is slowly fading.  English fluency was
associated with a very large wage premium early after the country
transitioned to a market economy, later its importance
declined but remained sizeable.

In contrast, the sole language that plays a role in determining
unemployment is Estonian.  Its fluency is associated with
approximately 5 percentage points lower probability of being unemployed.
These results suggest that Estonian labour market is      
segmented both along gender and the ethnic boundaries, men
more likely occupied in less communication-intense jobs.

The paper continues as follows: The next section provides a brief
background of Estonian society and labour market. In the following
section we discuss the data and provide some descriptive
statistics. In \cref{sec:method} we describe the econometric
approach, sections \ref{sec:results} and \ref{sec:extensions} present
the results for unemployment probability and wages.

@@

\section{Additional Evidence: Education and Workplace}
\label{sec:segregation_analysis}

@@

\label{sec:discussion}

>>>>

The bulk of the existing literature analyses the effect of knowledge
of the local majority language for immigrants.  It is well known that
language skills are associated with an income premium
between 5 and 35 percent
\citep{Chiswick1995,bleakley+chin2004,shields+price2002,leslie+lindley2001,chiswick+miller2002,Chiswick2010,Chiswick2015, Dustmann2003}.  For
instance, \citet{chiswick2008} argues that ``proficient'' speakers
typically earn about 15\% more than ``not proficient'' speakers.
The effect works largely through improving educational attainment
\citep{bleakley+chin2004,rooth+saarela2007native}.  Language skills are
also complementary to other types of human capital
\citep{chiswick+miller2007}.
Immigrants who are fluent in the local language also suffer
substantially less from unemployment \citep{shields+price2002, Dustmann2003}.

@@

There is also an increasing body of literature focusing on other languages than the local main business language.  These studies generally find
that English is a significant predictor of better income
\citep{Lang2009, Casale2011, Toomet2011, Williams2011, azam+2013EDandCC, isphording2013, fabo+2017E}.
The role of other languages, in particular in multi-lingual economy, is
less clear.  For instance,
\cite{Drinkwater1997} finds
a positive effect for Welsh speaking workers in
Wales and \cite{Armstrong2015} shows that bilingual native French-speakers enjoy
a significant premium both inside and outside of Quebec.
\citet{FrancoisGrin1998}, in contrast,
shows that Italian language skills in Swiss non-Italian speaking
regions are associated with lower wage.  It remains unclear if
this is a true effect or just an omitted variable bias.

Few previous studies have provided a comprehensive analysis of a labour
market from the viewpoint of language skills.
\citet{Chiswick1995, chiswick+miller2002, chiswick+miller2007,
  Bellante1998, Chiswick2010} analyse the effects of dominant language
fluency for earnings among immigrants in the United States, 
\citet{Chiswick1995} also do this for Canada, Australia and Israel.
\citet{leslie+lindley2001} analyse both employment and income but
focus on a single year and English language only.
\citet{YaoandOurs2015} only analyse the role of Dutch language
on labour market performance in Netherlands.

The literature on Central Eastern Europe (CEE) and former Soviet Union
countries is rather limited.  We know that the different development
trajectories have lead to disparities both favoring Russian-speaking
minority in Ukraine \citep{Constant2011} and hindering them in Estonia
\citep{Leping2008}.  \citet{Kahanec2009} show that non-citizens suffer
less favorable outcomes, in particular ethnic Russians in Estonia and
Latvia. 
\cite{Alan2015} show that Russian language proficiency is valuable in
the former USSR republic in Caucasus (Armenia, Azerbaijan and
Georgia) by increasing employment likelihood between six and nine
percentage points.  \citet{Toomet2011}, however, fails to find any
income premium from the majority language fluency in Estonia and
Latvia.  This outcome is corroborated by \citet{Lindemann2013} who shows that Estonian fluency
is associated with faster job finding rate while it does not
explain the lower occupational prestige of Russian speakers.  
A number of these studies have to overcome severe data limitations,
such as lack of fine-grained information on ethnic background in
EU-wide surveys \citep{Kahanec2010}.

In this paper, we analyse the relationship between both
majority and minority language skills, and English for both
wage and unemployment.  We analyse Estonia, a former Soviet republic
and now a rapidly growing and
transforming Easter European country.  It is characterized by a stark linguistic division
where the Russian-speaking community forms almost 30\% of
the population.  We use Estonian Labor Force Survey, one of the
higher-quality long-term datasets, very well suitable for our
purpose.  We provide long--term trends from the last years
of the planned economy in 1989 till the post-recession time in 2012.

We find that the Estonian language, the majority and the official
language of the country, is associated with very
small albeit slowly increasing wage premium over
the years.  Interestingly, this is true only for women, Russian men enjoy
virtually no premium from the majority language skills.
Russian skills, on the other hand, are clearly related to
higher wage but the relationship is slowly fading.  English fluency was
associated with a very large wage premium during the early years of
economic transition, later its importance
declined but remained sizeable.  
In contrast, the sole language that plays a role in determining
unemployment probability is Estonian.  Its fluency is associated with
approximately 5 percentage points lower probability of being unemployed.
These results suggest that Estonian labour market is      
segmented both along gender and the along ethnic background, with men
more likely occupied in less communication-intense jobs.

The paper continues as follows: The next section provides a brief
background of Estonian society and labour market. In the following
section we discuss the data and provide some descriptive
statistics. In \cref{sec:method} we describe the econometric
approach, sections \ref{sec:results} and \ref{sec:extensions} present
the results for unemployment probability and wages.  The two last
sections, \ref{sec:segregation-analysis} and \ref{sec:discussion}, are
devoted to additional analysis of labor market segregation, and
concluding discussion.

@@

\section{Additional Evidence: Education and Workplace}
\label{sec:segregation-analysis}

@@

\section{Discussion and Concluding Remarks}
\label{sec:discussion}
